Kindly Bent to Free Us

### 2019-11-22 kindly/lang.tex
Fixed typos.

In \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation and defer
the complete definition to \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
We then start from the leafs and grow the region
associated to each borrow until it can't grow anymore either due to a binding
or to conflicting uses. We demonstrate
on the following program:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
$a$ only has one borrow, its region covers its whole lexical scope.
$x$ has multiple conflicting borrow and requires more consideration.
We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal: making any region bigger would cause a borrowing
error.
Notes that indices of regions are well nested, as the inner region have
higher indices that the outer ones.

>>>>

In \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation and defer
the complete definition to \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
We then start from the leaves and grow the region
associated to each borrow until it can't grow anymore either due to a binding
or to conflicting uses. We demonstrate
on the following program:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
$a$ only has one borrow, its region covers its whole lexical scope.
$x$ has multiple conflicting borrows and requires more consideration.
We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal: making any region bigger would cause a borrowing
error.
Note that indices of regions are well nested, as the inner region have
higher indices than the outer ones.

### 2019-11-22 kindly/main.tex
conclusion
Add authors.
Adapt to the PLDI format.
Start improving the inference section.
moved session type example
formatting
Made a new pass on input parameters.
Gather the selected inference rules.
checkpoint - restructuring needed
Replace the pool example by an iterator one. Should be shorter.
Misc.
Metatheory: Introduced macro for the soundness theorem and moved it into 'metatheory-common.tex', as we want to state the theorem in both 'metatheory.tex' and 'metatheory-extra.tex'.
Reshuffle the intro again.
Metatheory: Used 'restatable' and removed metatheory-common again.
Fixed typos.

\documentclass[acmsmall,review,anonymous,table]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

% Math
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
% \usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage{textcomp}
\usepackage{gensymb}
% \usepackage{textgreek}
\usepackage{multicol}

@@

\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
%\setlist{leftmargin=*,noitemsep,topsep=5pt}

@@

@@

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

@@

  We present \lang, an extension of ML with constrained types that
  manages linearity and affinity properties through kinds. In addition
  \lang{} supports the exclusive and unrestricted borrowing of affine
  resources, inspired by features of Rust.
  Moreover, \lang{} retains the defining features of the ML family:
  an impure, strict functional expression language with complete principal type
  inference and type abstraction through modules.
  Our language does not require any linearity annotations in
  expressions and supports common functional programming idioms.
\end{abstract}

@@

\input{intro}
\input{motivation}
\input{lang}
\input{infer/infer}
\input{metatheory}
\input{soa}

@@

\clearpage
\appendix
\input{sdtyping/extra}
\clearpage
\input{infer/extra}
\clearpage
\input{lang-extra}
\input{metatheory-extra}

>>>>

\PassOptionsToPackage{svgnames}{xcolor}
\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

% Math
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{thm-restate}
% \usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage{textcomp}
\usepackage{gensymb}
% \usepackage{textgreek}
\usepackage{multicol}

@@

\usepackage{caption}
\usepackage{subcaption}
\usepackage[inline,shortlabels]{enumitem}
\setlist{leftmargin=*,noitemsep}
\usepackage{array}
\usepackage{bm}%Decent bolding for math symbols

@@

\mprset{sep=1.2em}

@@

\bibliographystyle{ACM-Reference-Format}

\author{Gabriel Radanne}
\affiliation{
  \institution{University of Freiburg}
  \country{Germany}
}
\email{radanne@informatik.uni-freiburg.de}

\author{Hannes Saffrich}
\affiliation{
  \institution{University of Freiburg}
  \country{Germany}
}
\email{saffrich@informatik.uni-freiburg.de}

\author{Peter Thiemann}
\affiliation{
  \institution{University of Freiburg}
  \country{Germany}
}
\email{thiemann@informatik.uni-freiburg.de}

@@

  We present \lang, an extension of ML with constrained types that
  manages linearity and affinity properties through kinds. In addition
  \lang{} supports the exclusive and shared borrowing of affine
  resources, inspired by features of Rust.
  Moreover, \lang{} retains the defining features of the ML family:
  an impure, strict, functional expression language with complete principal type
  inference and type abstraction through modules.
  Our language does not require any linearity annotations in
  expressions and supports common functional programming idioms.
\end{abstract}

@@

\input{intro}
\input{motivation}
\input{lang}
\input{infer/infer}
\input{metatheory}
\input{soa}
\input{conclusion}

@@

\clearpage
\appendix

\section{Further Examples}
\label{sec:extra-example}
\lstMakeShortInline[keepspaces,basicstyle=\small\ttfamily]@
\input{motivation-linearity}
\input{motivation-persistent}
\input{motivation-pool}
\lstDeleteShortInline@

\input{sdtyping/extra}
\clearpage
\input{infer/extra}
\clearpage
\input{lang-extra}
\input{metatheory-extra}

### 2019-11-22 kindly/motivation-persistent.tex
Fixed typos.

\cref{sig:hybarray} contains an API @HybArray@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HybArray@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}[tp]
  \centering
  % \begin{subfigure}{1\linewidth}
\begin{lstlisting}
module type HybArray = sig
  include ARRAY
  val set : &('a t) * int * 'a -> 'a t
  val set_mut : &!('a t) * int * 'a -> unit
end

module Cow : HybArray = struct
  include Array
  let set (a, i0, x0) =
    Array.mapi (fun i x -> if i = i0 then x0 else x) a
  let set_mut = Array.set
end
\end{lstlisting}
    \vspace{-15pt}
  %   \caption{Implementation using copy-on-write}
  % \end{subfigure}
  \caption{Hybrid arrays}
    \label{sig:hybarray}
    \label{ex:cow}

>>>>

\cref{sig:hybarray} contains an API @HYBARRAY@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HYBARRAY@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}[tp]
  \centering
  % \begin{subfigure}{1\linewidth}
\begin{lstlisting}
module type HYBARRAY = sig
  include ARRAY
  val set : &('a t) * int * 'a -> 'a t
  val set_mut : &!('a t) * int * 'a -> unit
end

module Cow : HYBARRAY = struct
  include Array
  let set (a, i0, x0) =
    Array.mapi (fun i x -> if i = i0 then x0 else x) a
  let set_mut = Array.set
end
\end{lstlisting}
    \vspace{-15pt}
  %   \caption{Implementation using copy-on-write}
  % \end{subfigure}
  \caption{Hybrid arrays}
    \label{sig:hybarray}
    \label{ex:cow}
